---
title: "Fundamentals of Linear Models 2025-2026 - Assignment #1"
date: "16 September 2025"
output:
  pdf_document: default
  html_notebook: default
---

## Introduction to the Assignment

This document you are reading is an [R Markdown](http://rmarkdown.rstudio.com) Notebook. When you execute code within the notebook, the results appear beneath the code. It is a very efficient way to document and maintain your work including useful anotations, etc. If you want to learn more about R Markdown, read and follow the available [online tutorials](https://rmarkdown.rstudio.com/lesson-1.html)

## Task#1: Creating a dataframe

------------------------------------------------------------------------

Create a dataframe coming up with a sample data of your own (possibly made up) with the following contents:

-   4 columns/variables: one column with a factor with two levels, another column with a factor with 3 levels and two columns with numeric values.

-   6 rows or observations

```{r}



```

## Task#2: loading required libraries

------------------------------------------------------------------------

Read through the complete assignment and identify which libraries will be required for the tasks and load them

```{r}
# add code to load the required libraries



```

## Task#3: Working with and exploring dataframes

------------------------------------------------------------------------

From the *`languageR`* library we will use a sample dataframe called `lexdec`. This dataset contains lexical decision latencies elicited from 21 subjects for 79 English concrete nouns. To understand the content of each variable, look at the description of the dataset using `?lexdec` as described during the workgroup.

1.  **Show the first 4 rows of the dataframe**

    *Tip*: you can use the `head()` function. To learn how to use it, type `help(head))` or `?head` in the terminal.

```{r}


```

2.  **Extract the column names from the dataframe.**

```{r}


```

3.  **Sort the dataframe by the reaction time** (`RT` variable) **in ascending order and then in descending order**

```{r}


```

5.  **Create a new dataframe called** `lexdec_reduced`**with only the variables from** `Subject` **to** `Frequency` **and selecting only the entries from native English speakers (as coded in the** `NativeLanguage` **variable).**

```{r}


```

6.  **Save** `lexdec_reduced` **dataframe in a file in the** `/data` **directorate**.

    Note: Choose any name and data format that you want for the file.

```{r}


```

## Task#4 (Extra Credit)

------------------------------------------------------------------------
