---
editor: visual
---

# Simple Linear Model reporting {.unnumbered}

A report of the results of linear models should include the following elements:

-   **Model description**

    -   Descriptive sentence on the modeled data and predictor: *A* *simple linear regression analysis was used to test if \[Outcome Variable\] significantly predicted \[Predictor Variable\]*

    -   Model summary overall fit including *R2* and *F-test* results. *The model explained ...% of the variance in \[Outcome variable\], F(df1, df2)=..., p\<..., R2 =..."*

-   **Individual predictors:**

    -   For each predictor report: coefficient value (also called *unstandardized beta*[^model_reporting1-1]), the Confidence Interval, t-statistic and p-value.

        e.g.: "*It was found that* *\[Predictor variable\] significantly predicted \[response variable\] (*$\beta=..., CI=[..,..],t=...,p=...)$.

-   **Assumptions check:** explain how you checked assumptions, including if any data transformation was required, etc.

[^model_reporting1-1]: We will discussed standardized coefficient values in the next workgroups on multiple regression.

Let's look at an example based again on the example on assignment 3:

```{r}
#| warning: false
library(tidyverse)

dfVowel<-read.csv('./data/extended_vowel_duration.csv')

null_model <- lm(Duration~1,data=dfVowel)
mod <- lm(Duration~logFreqs, data = dfVowel)
summary(mod)
```

The result contains all the information required, with the exception of the Confidence Interval for the logFreqs coefficient.

This can be conveniently extracted using the function `confint()` that takes a model as a parameter.

```{r}
confint(mod)
```

The reporting of the analysis could be written as follows:

::: callout-note
Simple linear regression was performed to assess the relationship between *vowel production Duration* and the logarithmically transformed *frequency* of the word produced. Assumption checks revealed no violations in normality and homoscedasticity, as indicated by the residual plots and non significant results from the Shapiro-Wilk test.

After transformation, the fitted regression model was:

$$
Duration = b_0+b_1 \times Frequency
$$

The model explained $55\%$ of the variance in vowel duration ( $F(1,598)=449.30$ $,p<.001,R^2=0.45$). It was found that the logarithmic transformed word Frequency significantly predicted the vowel duration ($\beta = -5.64$, $95\%CI[-6.13,-5.14]$, $t=136.52, p<.001$).
:::
