---
editor: "visual"
date: 08/12/2025
bibliography: references.bib
execute:
  echo: true
  warning: false
  message: false
---

# Multicollinearity {.unnumbered}

Multicollinearity occurs when predictors in a regression model are highly correlated. This is common in variables as for example Frequency, Word Length or Predictability (shorted words are less frequent and more predictable than long words).

Fitting a model with two predictors that vary together creates problems

-   Inflated standard errors
-   Unstable coefficient signs
-   Difficulty interpreting effects
-   Lower statistical power

Let's examine examine relationships among using the `lexdec` dataset Frequency, FamilySize, WordLength, and logRT.

```{r}
library(languageR)
data(lexdec)
```

## Correlation assessment using VIF

Variance Inflation Factor (VIF) quantifies how much variance of a coefficient is inflated due to correlation. The VIF is calculated trying to create a regression against other predictors **VIF = 1 / (1 - R²)**.

-   A high R² for a predictor means it's predictable by others (high multicollinearity), which inflates its VIF, making its coefficient unstable. 

Interpretation guidelines:

-   **VIF = 1:** No multicollinearity; the predictor is independent.

-   **1 \< VIF \< 5:** Moderate multicollinearity; generally not a serious problem.

-   **VIF \> 5:** Potentially severe multicollinearity; coefficients and p-values may be unreliable, warranting investigation.

-   **VIF \> 10:** Serious multicollinearity; the model is likely unstable, and the predictor might need removal or transformation. 

We calculate it with the `vif()` function in the `car` model.

```{r}
library(car)

mod1 <- lm(RT ~ Frequency + FamilySize + Length, data = lexdec)

vif(mod1)
```

------------------------------------------------------------------------

## Multicollinearity in Interaction Models

Interaction terms are inherently correlated with their main components, and values would be higher.

```{r}
mod2<-lm(RT~FamilySize+Frequency*Length, data=lexdec)

vif(mod2)
```

As the interactions are introduced, the variance inflation increases. In those cases, it is recommended to center all continuous variables. Centering removes the impact on the variance.

```{r}
lexdec$Frequency_c <- lexdec$Frequency - mean(lexdec$Frequency)
lexdec$Length_c <- lexdec$Length - mean(lexdec$Length)

mod3 <- lm(RT ~ Frequency_c * Length_c, data = lexdec)
vif(mod3)
```

------------------------------------------------------------------------

# Summary of best practices

-   **Center continuous predictors** before interaction creation
-   **Sum-code categorical variables**
-   Report modeling decisions clearly

------------------------------------------------------------------------

# Reporting Example

> Predictors were centered before computing the interaction. VIF values for main effects were below 2, and the interaction term showed also a low VIF = 1.01.
