---
title: "Workgroup 2: Data Exploration with R"
author: "Leticia Pablos Robles"
---

# ðŸ§  Learning Objectives

By the end of this lesson, you will be able to:

-   Import and explore datasets.
-   Use basic functions and create simple plots.

# ðŸ“‚ Dataset loading and importing in R

## Using data from a package
```{r}
library(languageR)

```


## Importing data from a text file
```{r}
#data <- read.csv("yourfile.csv")  # Load a CSV file  
#head(data)                        # View the first few rows 
```


``` r

```

## Importing data from an Excel file

# Dataset exploration

## Understanding data structure

## Defining factors

# Data organization and transformation

## Sorting, filtering, merging

## Data transformation

## Using *tidyverse* library

# ðŸ“ Exercises

## Explore a Built-in Dataset

-Load the library called ***languageR**,* using the following command *.* In the languageR library, there are some sample datasets that we will use for future workgroups.

``` r
```

`{r} # load required libraries library(languageR) library(tidyverse)}`

From the *languageR* library we will use a sample dataframe called *spanishMeta*. This dataset is based on a more general dataset called *spanish* also included in the library, from *Spassova, M. S. (2006) Las marcas sintacticas de atribucion forense de autoria de textos escritos en espanol, Masters thesis, Institut Universitari de Linguistica Aplicada, Universitat Pompeu Fabra, Barcelona*

### Display dataframe

To view the contents of a dataset we can simply write the name of the dataset in the console

`{r} spanishMeta}`

If we have a lot of data, the console display will be very long. Then you can use the *head()* function to only see the first few rows of the dataset

`{r} head(spanishMeta)}`

By default the function shows the first 6 rows of the dataset. if you look at the options of the function using *help(head)*, you can see that there is a parameter (called argument in the vocabulary of R) *n* that can be used to indicate the number of rows to show. For the exercise, to show 4 rows, you can use:

`{r} head(spanishMeta, n=4)}`

### Extract column names

There a number of functions that allow to inspect data sets: \* *str()*: provides a brief description of the data set variables and contents \* *names()*: lists the names of the variables \* *summary()*: gives basic summary statistics for every variable (we will explore this in the next assignment) \* *head()*: shows the first rows of the dataset (as we saw above) \* *tail()*: shows the last rows of the dataset

In the assignment it was asked to list the names of the variables in the dataset. The easiest way is to use the *names()* function

`{r} names(spanishMeta)}`

### Extract vector

Accessing the variables of a dataset is possible using the *\$* operator. So for example, to list the contents of the *PubDate*

`{r} spanishMeta$PubDate}`

In the exercise it was asked to take this information and include it in a variable called *PublicationDate*. Using the above notation this is simply done as

`{r} PublicationDate<-spanishMeta$PubDate PublicationDate}`

### Sort a vector

You can sort the vector using the *sort()* function as below.

`{r} sort(PublicationDate)}`

`{r} sort(PublicationDate,decreasing = TRUE)}`

Finally, there are many functions available to work, modify, and access dataframes. As an example let's see how we could select a part of the dataframe based on a criteria. If we want to select the entries that were published before 1980 for example, we can do the following

`{r} spanishMeta[spanishMeta$PubDate<1980,]}`

You can also use the *subset()* function to make a selection

`{r} subset(spanishMeta, PubDate<1980)}`

For all the operations above you can also use pipes:

`{r} library(dplyr)}`

`{r} PublicationDate %>% sort}`

`{r} PublicationDate %>% sort(decreasing = TRUE)}`

`{r} spanishMeta %>% filter(PubDate<1980)}`

For self-study and exploration look at [this website](https://www.statmethods.net/management/index.html) to get information on basic manipulation of dataframes.

## 
